Spá: a web-based viewer for text mining in Evidence Based Medicine

### 2014-05-01 documentation/paper.tex
Address some comments on the paper

% Created 2014-04-14 Mon 18:10
\documentclass[runningheads,a4paper]{llncs}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage[multiple]{footmisc}
\usepackage{acronym}
\tolerance=1000
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{listings}

@@

Summarizing the evidence about medical interventions is an immense undertaking, in part because unstructured \ac{pdf} documents remain the main vehicle for disseminating scientific findings.
Clinicians and researchers must therefore manually extract and synthesise information from these \acp{pdf} to be published in \emph{systematic reviews}.
We introduce Spá\footnote{From the Old Norse word spá or spæ referring to prophesying (prophecy)}\footnote{Source code available under GPLv3 at \url{https://github.com/joelkuiper/spa} \cite{Kuiper2014}; demo available at \url{http://spa.clinici.co/}}, a web-based viewer that enables automated annotation and summarisation of \acp{pdf} via \ac{ml}.
To illustrate its functionalility, we use Spá to semi-automate the assessment of bias in clinical trials.
%Spá can visualize the output from hybrid models that simultaneously classify documents (e.g., identifying trials as low or high risk of various biases), and annotate sentences that support these classifications.
Spá has a modular architecture, therefore the tool may be widely useful in other domains with a \ac{pdf}-based literature, including law, physics, and biology.

@@

Imposing structure on full-text documents is an important and practical task in natural language processing and machine learning.
\emph{Systematic reviews} are an instructive example.
Such reviews aim to answer clinical questions by providing an exhaustive synthesis of all the current evidence in published literature.
They are fundamental tools in \ac{ebm} \cite{Sackett1996,Valkenhoef2012}.
To produce these reviews data must be manually extracted from the literature.
These extraction tasks are extremely labourous, but could potentially be (semi-)automated using machine learning approaches.

As an example we consider risk of bias assessment, here reviewers assess, e.g., whether study participants and personnel were properly blinded \cite{Higgins2011}.
Assessing risk of bias is a time-consuming task.
A single trial may typically take a domain expert ten minutes \cite{Hartling2011}, and a single review may typically include several dozen trials.
Making matters worse, due to low rates of reviewer agreement it is regarded as best practice to have each study assessed twice by independent reviewers who later come to a consensus \cite{Hartling2009}.

@@

While our application of interest is \ac{ebm}, we emphasize that the visualization tool can be used for any domain in which one wants to annotate \acp{pdf}.
Thus the contribution of this work is two-fold, as we present:
(1) a practical tool that incorporates machine learning to help researchers rapidly assess the risk of biases in published biomedical articles, and,
(2) a general open-source web tool for visualizing the predictions of trained models from full-text articles.
%These contributions are described further in sections \ref{section:EBM-ML} and \ref{section:architecture}.

\section{Automating \acl{ebm}}
\label{section:EBM-ML}

@@

For each of the annotations the relevant nodes in the document are highlighted.
A custom scrollbar, inspired by \href{http://substance.io/}{substance.io} that acts as a `mini-map' is projected to show where annotations reside within the document.
The user can interactively activate and inspect specific results.

@@

More generally, we believe the tool to be potentially useful for a much wider range of \acl{ml} applications.
To increase the generality of the tool we are developing a pluggable system for processing topologies, allowing developers to quickly plug in new systems for automated \ac{pdf} annotation.
Furthermore, we are working to allow users to save selected annotations, possibly embedded within the document itself, for sharing and off-line use.
The vision is to have an extensible system for machine assisted data extraction and data summarization.

>>>>

% Created 2014-04-14 Mon 18:10
\documentclass[runningheads,a4paper]{llncs}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage[multiple]{footmisc}
\usepackage{acronym}
\tolerance=1000
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{listings}

@@

Summarizing the evidence about medical interventions is an immense undertaking, in part because unstructured \ac{pdf} documents remain the main vehicle for disseminating scientific findings.
Clinicians and researchers must therefore manually extract and synthesise information from these \acp{pdf} to be published in \emph{systematic reviews}.
We introduce Spá\footnote{From the Old Norse word spá or spæ referring to prophesying (prophecy)}\footnote{Source code available under GPLv3 at \url{https://github.com/joelkuiper/spa} \cite{Kuiper2014}; demo available at \url{http://spa.clinici.co/}}, a web-based viewer that enables automated annotation and summarisation of \acp{pdf} via \acl{ml}.
To illustrate its functionality, we use Spá to semi-automate the assessment of bias in clinical trials.
%Spá can visualize the output from hybrid models that simultaneously classify documents (e.g., identifying trials as low or high risk of various biases), and annotate sentences that support these classifications.
Spá has a modular architecture, therefore the tool may be widely useful in other domains with a \ac{pdf}-based literature, including law, physics, and biology.

@@

Imposing structure on full-text documents is an important and practical task in natural language processing and machine learning.
\emph{Systematic reviews} are an instructive example.
Such reviews aim to answer clinical questions by providing an exhaustive synthesis of all the current evidence in published literature.
They are fundamental tools in \ac{ebm} \cite{Sackett1996,Valkenhoef2012}.
Data must be manually extracted from the literature to produce the systematic reviews.
These extraction tasks are extremely laborious, but could potentially be assisted by machine learning approaches.

As an example we consider risk of bias assessment.
Here reviewers assess, e.g., whether study participants and personnel were properly blinded \cite{Higgins2011}.
Assessing risk of bias is a time-consuming task.
A single trial may typically take a domain expert ten minutes \cite{Hartling2011}, and a single review may typically include several dozen trials.
Making matters worse, due to low rates of reviewer agreement it is regarded as best practice to have each study assessed twice by independent reviewers who later come to a consensus \cite{Hartling2009}.

@@

While our application of interest is \ac{ebm}, we emphasize that the visualization tool can be used for any domain in which one wants to annotate \acp{pdf}, e.g. genome-wide association studies or jurisprudence.
Thus the contribution of this work is two-fold, as we present:
(1) a practical tool that incorporates machine learning to help researchers rapidly assess the risk of biases in published biomedical articles, and,
(2) a general open-source web tool for visualizing the predictions of trained models from full-text articles.
%These contributions are described further in sections \ref{section:EBM-ML} and \ref{section:architecture}.

\section{Case study: Risk of Bias in \acl{ebm}}
\label{section:EBM-ML}

@@

For each of the annotations the relevant nodes in the document are highlighted.
A custom scrollbar, inspired by \href{http://substance.io/}{substance.io}, that acts as a `mini-map' is projected to show where annotations reside within the document.
The user can interactively activate and inspect specific results.

@@

We believe the tool to be potentially useful for a much wider range of \acl{ml} applications.
Currently we are developing a pluggable system for processing topologies, allowing developers to quickly plug in new systems for automated \ac{pdf} annotation.
Furthermore, we are working to allow users to persist annotations and marginalia, possibly embedded within the document itself, for sharing and off-line use.
The vision is to have an extensible system for machine assisted data extraction that will greatly increase both the quality and the reproducibility (i.e. data provenance) of current \acl{ebm}.
